Location and Scale-Invariant Power Transformations for Transforming Data to Normality

### 2025-04-11 manuscript_latex/manuscript.tex
Update references.

Many statistical and some machine learning methods assume normality of
the underlying data, e.g.~analysis of variance and linear discriminant
analysis. However, numerical features in datasets may strongly deviate
from normal distributions, e.g.~by being skewed. Power transformations
aim to stabilise variance and improve normality of such features
(Bartlett 1947; John W. Tukey 1957). The two most commonly used
transformations are that of Box and Cox (1964) and Yeo and Johnson
(2000).

@@

Real-world data may contain outliers, to which maximum likelihood
estimation can be sensitive. Their presence may lead to poor
transformations to normality, as shown in Figure
\ref{fig:decreased-normality}. As indicated by Raymaekers and Rousseeuw
(2024), the general aim of power transformations should be to transform
non-outlier data to normality, i.e.~achieve \emph{central normality}. To
achieve this, they devised an iterative procedure to find a robust
estimate of the transformation parameter \(\lambda\). Briefly, this
process requires identifying outliers in the data and weighting such
instances during the optimisation process. Raymaekers and Rousseeuw
(2024) achieve this through weighted maximum likelihood estimation.
However, because this procedure iteratively estimates and updates
\(\lambda\), it can not be used here to simultaneously estimate
\(\lambda\), \(x_0\) and \(s\) for location- and scale-invariant power
transformations. Nonetheless, as a procedure, weighted MLE can be used
for estimating the transformation, shift and scale parameters.

Here, weighted maximum likelihood estimation is based on equations
\ref{eqn:box-cox-invariant-log-likelihood} and
\ref{eqn:yeo-johnson-invariant-log-likelihood}. Compared to Raymaekers
and Rousseeuw (2024), these log-likelihood functions include additional
terms to accommodate estimation of \(x_0\) and \(s\). The weighted
location- and scale-invariant Box-Cox log-likelihood function is:

@@

\begin{itemize}
\item
  A step function, with \(\delta_1 \geq 0\) as threshold parameter:
  \begin{equation*}
  w_i =
  \begin{cases}
  1 & \text{if } \left| \dot{x}_i \right| \leq \delta_1\\
  0 & \text{if } \left| \dot{x}_i \right| > \delta_1
  \end{cases}
  \end{equation*}
\item
  A triangle function (or generalised Huber weight), with
  \(\delta_1 \geq 0\) and \(\delta_2 \geq \delta_1\) as threshold
  parameters: \begin{equation*}
  w_i =
  \begin{cases}
  1 & \text{if } \left| \dot{x}_i \right| < \delta_1\\
  1 - \frac{\left| \dot{x}_i \right| - \delta_1}{\delta_2 - \delta_1} & \text{if } \delta_1 \leq \left| \dot{x}_i \right| \leq \delta_2 \\
  0 & \text{if } \left| \dot{x}_i \right| > \delta_2
  \end{cases}
  \end{equation*}
\item
  A tapered cosine function (J. W. Tukey 1967), with \(\delta_1 \geq 0\)
  and \(\delta_2 \geq \delta_1\) as threshold parameters:
  \begin{equation*}
  w_i =
  \begin{cases}
  1 & \text{if } \left| \dot{x}_i \right| < \delta_1\\
  0.5 + 0.5 \cos\left(\pi \frac{\left| \dot{x}_i \right| - \delta_1}{\delta_2 - \delta_1} \right) & \text{if } \delta_1 \leq \left| \dot{x}_i \right| \leq \delta_2 \\
  0 & \text{if } \left| \dot{x}_i \right| > \delta_2
  \end{cases}
  \end{equation*}
\end{itemize}

@@

We require a quantile function (or an approximation thereof) to draw
random values from an asymmetric generalised normal distribution using
inverse transform sampling. Gijbels et al.~derived the quantile function
\(F_{\alpha}^{-1}(p)\), which incorporates the quantile function of the
symmetric generalised normal distribution derived by Griffin (Griffin
2018):

@@

Power transformations aim to transform features to a normal
distribution. However, this may not always be successful or possible.
Deviations from normality can be detected by normality tests, such as
the Shapiro-Wilk test (Shapiro and Wilk 1965). In practice, normality
tests may be too stringent with large sample sizes, outliers, or both.
Here we develop a test for central normality.

@@

Let the residual error for each element of $\mathbf{X}$ be $r_i =\left| \frac{x_i - \mu_M}{\sigma_M} - F^{-1}_{\mathcal{N}}(p_i) \right|$, 
with $\mu_M$ and $\sigma_M$ robust Huber M-estimates of location and scale of $\mathbf{X}$ [@Huber1981-su], 
and $F^{-1}_{\mathcal{N}}$ the quantile function of the normal distribution $\mathcal{N}(0, 1)$.

@@

We used simulated data to assess invariance to location and scale of the
proposed power transformations, to develop the empirical central
normality test, and to determine weighting for robust transformations.
The \(\lambda\) parameter for conventional power transformations (Eqn.
\ref{eqn:box-cox-original} and \ref{eqn:yeo-johnson-original}), as well
as \(\lambda\), \(x_0\) and \(s\) parameters for location- and
scale-invariant power transformations (Eqn. \ref{eqn:box-cox-invariant}
and \ref{eqn:yeo-johnson-invariant}) were estimated using the BOBYQA
algorithm for derivative-free bound constraint optimisation (Powell
2009) through maximum likelihood estimation. The required algorithms
were implemented in the \texttt{power.transform} R software package
(Zwanenburg and L\"ock 2024b) (version 1.0.1). Of note, the
\texttt{power.transform} package shifts feature values into the positive
domain if negative or zero values are present for Box-Cox power
transformations.

@@

Outlier values were then drawn to randomly replace 10 percent of the
elements of \(\mathbf{X}_i\). Outlier values were set according to John
W. Tukey (1977), as follows. Let \(x^{*} \sim U\left(-2, 2\right)\).
Then the corresponding outlier value was:

@@

In addition to no power transformation and location- and scale-invariant
power transformations, conventional and Raymaekers and Rousseeuw's
robust adaptation (Raymaekers and Rousseeuw 2024) were assessed. For the
latter two methods, normalisation before standardisation using was
additionally assessed using the following two methods:

@@

A common feature in health-related datasets is age. Here we use data on
228 patients with lung cancer that was collected and published by
Loprinzi et al. (Loprinzi et al. 1994). The age in the cohort was
\(62.4 \pm 9.1\) (mean ± standard deviation) years. Applying
conventional and invariant Yeo-Johnson transformations to patient age
yielded the following results, see Figure
\ref{fig:experimental-results-invariance}: no transformation
(\(\sum r_i = 16.5\), \(p = 0.69\)); conventional transformation
(\(\lambda = 2.0\), \(\sum r_i = 11.5\), \(\mu_{YJ} = 1.8 \cdot 10^3\),
\(\sigma_{YJ} = 0.5 \cdot 10^3\), \(p = 0.96\)); Raymaekers and
Rousseeuw's robust adaptation (\(\lambda = 2.0\), \(\sum r_i = 11.5\),
\(\mu_{YJ} = 1.8 \cdot 10^3\), \(\sigma_{YJ} = 0.5 \cdot 10^3\),
\(p = 0.96\)); location- and scale-invariant transformation
(\(\lambda = 1.3\), \(\sum r_i = 8.8\), \(\mu_{YJ} = -1.2\),
\(\sigma_{YJ} = 1.1\), \(p = 0.98\)); and robust location- and
scale-invariant transformation (\(\lambda = 1.3\), \(\sum r_i = 9.3\),
\(\mu_{YJ} = -1.0\), \(\sigma_{YJ} = 1.1\), \(p = 0.93\)).

@@

Gorman, Williams and Fraser recorded body mass (in grams) of 342
penguins of three different species (Gorman, Williams, and Fraser 2014).
The body mass was \((4.2 \pm 0.8) \cdot 10^3\) (mean ± standard
deviation) grams, and not centrally normal (\(p < 0.001\)). Applying
conventional and invariant Yeo-Johnson transformations to body mass
yielded the following results, see Figure
\ref{fig:experimental-results-invariance}: no transformation (residual
sum \(\sum r_i = 48.0\), \(p < 0.001\)); conventional transformation
(\(\lambda = -0.5\), \(\sum r_i = 32.2\), \(\mu_{YJ} = 2.1\),
\(\sigma_{YJ} = 4 \cdot 10^{-3}\), \(p = 0.10\)); Raymaekers and
Rousseeuw's robust adaptation (\(\lambda = -0.5\), \(\sum r_i = 32.2\),
\(\mu_{YJ} = 2.1\), \(\sigma_{YJ} = 4 \cdot 10^{-3}\), \(p = 0.10\));
location- and scale-invariant transformation (\(\lambda = 0.5\),
\(\sum r_i = 26.8\), \(\mu_{YJ} = 0.9\), \(\sigma_{YJ} = 0.9\),
\(p = 0.28\)); and robust location- and scale-invariant transformation
(\(\lambda = 0.3\), \(\sum r_i = 22.0\), \(\mu_{YJ} = 0.7\),
\(\sigma_{YJ} = 0.9\), \(p = 0.69\)).

@@

Geospatial datasets usually contain coordinates. The Ames housing
dataset contains data on 2930 properties that were sold between 2006 and
2010 (De Cock 2011), including their geospatial coordinates. The
latitude was \(42.03 \pm 0.02)\) (mean ± standard deviation). Applying
conventional and invariant Yeo-Johnson transformations to latitude
yielded the following results, see Figure
\ref{fig:experimental-results-invariance}: no transformation (residual
sum \(\sum r_i = 328\), \(p < 0.001\)); conventional transformation
(\(\lambda = 62.1\), \(\sum r_i = 319\),
\(\mu_{YJ} = 4.8 \cdot 10^{99}\), \(\sigma_{YJ} = 0.1 \cdot 10^{99}\),
\(p < 0.001\)); Raymaekers and Rousseeuw's robust adaptation
(\(\lambda = 95.4\), \(\sum r_i = 315\),
\(\mu_{YJ} = 6.4 \cdot 10^{153}\), \(\sigma_{YJ} = 0.3 \cdot 10^{153}\),
\(p < 0.001\)); location- and scale-invariant transformation
(\(\lambda = 1.5\), \(\sum r_i = 326\), \(\mu_{YJ} = -1.2\),
\(\sigma_{YJ} = 0.8\), \(p < 0.001\)); and robust location- and
scale-invariant transformation (\(\lambda = 1.1\), \(\sum r_i = 308\),
\(\mu_{YJ} = -1.3\), \(\sigma_{YJ} = 1.2\), \(p < 0.001\)).

@@

The Top Gear dataset contains data on 297 vehicles that appeared on the
BBC television show \emph{Top Gear} (Alfons 2021). Within this dataset,
the fuel consumption feature contains outliers due to highly
fuel-efficient vehicles. Applying conventional and invariant Yeo-Johnson
transformations to the fuel consumption feature yielded the following
results, see Figure \ref{fig:experimental-results-outlier-robustness}:
no transformation (residual sum \(\sum r_i = 54\), \(p=0.72\));
conventional transformation (\(\lambda = -0.1\), \(\sum r_i = 55\),
\(\mu_{YJ} = 3.0\), \(\sigma_{YJ} = 0.3\), \(p < 0.001\)); Raymaekers
and Rousseeuw's robust adaptation (\(\lambda = 0.8\), \(\sum r_i = 48\),
\(\mu_{YJ} = 29\), \(\sigma_{YJ} = 15\), \(p=0.37\)); location- and
scale-invariant transformation (\(\lambda = -1.3\), \(\sum r_i = 44\),
\(\mu_{YJ} = 0.5\), \(\sigma_{YJ} = 0.1\), \(p < 0.001\)); and robust
location- and scale-invariant transformation (\(\lambda = 1.0\),
\(\sum r_i = 56\), \(\mu_{YJ} = 1.7\), \(\sigma_{YJ} = 1.0\),
\(p=0.76\)).

@@

The ischemic stroke dataset contains historic data from 126 patients
with risk at ischemic stroke (Kuhn and Johnson 2019). These patients
underwent Computed Tomography Angiography to characterize the carotid
artery blockages. Angiography imaging was then assessed, and various
characteristics related to the blood vessels and the disease are
measured. The maximum arterial wall thickness feature contains several
instances with outlier values. Applying conventional and invariant
Yeo-Johnson transformations to this feature yielded the following
results, see Figure \ref{fig:experimental-results-outlier-robustness}:
no transformation (residual sum \(\sum r_i = 110\), \(p=0.83\));
conventional transformation (\(\lambda = -0.7\), \(\sum r_i = 30\),
\(\mu_{YJ} = 1.0\), \(\sigma_{YJ} = 0.1\), \(p=0.003\)); Raymaekers and
Rousseeuw's robust adaptation (\(\lambda = 1.1\), \(\sum r_i = 136\),
\(\mu_{YJ} = 7.2\), \(\sigma_{YJ} = 14.3\), \(p=0.88\)); location- and
scale-invariant transformation (\(\lambda = 0.2\), \(\sum r_i = 12\),
\(\mu_{YJ} = -11.8\), \(\sigma_{YJ} = 6.9\), \(p=0.15\)); and robust
location- and scale-invariant transformation (\(\lambda = -0.3\),
\(\sum r_i = 30\), \(\mu_{YJ} = 0.8\), \(\sigma_{YJ} = 0.2\),
\(p=0.18\)).

@@

We used 231 datasets containing at least one numeric feature from the
Penn Machine Learning Benchmarks collection (Romano et al. 2022). In
this collection, 114 datasets correspond to regression tasks and 117
datasets to classification tasks. Using the familiar auto-machine
learning library (Zwanenburg and L\"ock 2024a) (version 1.5.0), each
dataset was used to train a model for each of 32 process configurations.
Each process configuration specifies the learner (generalised linear
model, L1-regularised linear models (Lasso), gradient boosted linear
model, or random forest), transformation method (none, conventional
Yeo-Johnson, robust invariant Yeo-Johnson, robust invariant Yeo-Johnson
with empirical central normality test (rejecting transformations with
\(p \leq 0.01\)), and normalisation method (none,
\(z\)-standardisation), yielding 32 distinct configurations. Before each
experiment, each dataset was randomly split into a training (70\%) and
holdout test (30\%) set five times. Thus, a total of 36960 models were
created. Each model was then evaluated using the holdout test set using
one of two metrics, i.e.~the root relative squared error (RRSE) for
regression tasks and the area under the receiver operating
characteristic curve (AUC) for classification tasks.

@@

To estimate the marginal effects of process parameters, including
transformation method, we first fit a regression random forest (ranger
package (Wright and Ziegler 2017) version 0.16.0): 2000 trees, node size
2, other hyperparameters default) with process parameters and task
difficulty as predictors and normalised rank as response variable. The
estimated marginal effects are shown in Figure
\ref{fig:marginal-effect-plot}. On the scale of normalised ranks
(\([0.0, 1.0]\)), only the random forest performed better than the
average of all learners (rank difference \(0.173\)). The marginal
improvement in performance from z-standardisation was \(0.011\).
Transformation methods had the following marginal effects: \(-0.001\)
for using conventional Yeo-Johnson transformation instead of no
transformation; \(-0.012\) for using robust invariant Yeo-Johnson
transformation instead of no transformation; \(-0.012\) for using robust
invariant Yeo-Johnson transformation with empirical central normality
test instead of no transformation; \(-0.012\) for using robust invariant
Yeo-Johnson transformation instead of conventional Yeo-Johnson
transformation; and \(-0.012\) for using using robust invariant
Yeo-Johnson transformation with empirical central normality test instead
of conventional Yeo-Johnson transformation.

@@

In their work on power transformation, Box and Cox already mention
transformation with a shift parameter, but preferred the version in Eq.
\ref{eqn:box-cox-original} for the theoretical analysis in their paper
(Box and Cox 1964), which subsequently became the convention. Yeo and
Johnson's power transformation lacks a shift parameter altogether (Yeo
and Johnson 2000). We showed that these power transformations are
sensitive to location and scale of data distributions. To mitigate this
issue, we defined location- and scale-invariant variants of the Box-Cox
and Yeo-Johnson transformations. We furthermore assessed methods for
making these transformations robust to outliers, and devised an
empirical test for central normality.

@@

We introduced an empirical test for central normality to assess whether
sequences deviate from normality in a way that might require closer
inspection prior to further processing. The empirical test for central
normality differs from other tests for normality, such as the
Shapiro-Wilk test (Shapiro and Wilk 1965), as it assesses normality of
the central portion of a feature, instead of the entire feature.
Compared to the central normality test and the Shapiro-Wilk test, the
empirical central normality test remains consistent in the presence of
outliers, although the former tests are more powerful when outliers are
absent.

@@

Below we will focus on criteria based on the empirical density function
and those based on skewness and kurtosis of the transformed featured.
Other potential criteria, such as the Shapiro-Wilk test statistic
(Shapiro and Wilk 1965) are not investigated here. In the case of the
Shapiro-Wilk test statistic this is because of lack of scalability to
features with many (\(> 5000\)) instances, and because adapting the test
statistic to include weights is not straightforward.

@@

The first class of criteria is based on the empirical distribution
function (EDF). Transformation parameters are then fit through
minimisation of the distance between the empirical distribution function
\(F_{\epsilon}\) and the cumulative density function (CDF) of the normal
distribution \(F_{\mathcal{N}}\). Let
\(F_{\epsilon}\left(x_i \right) = \frac{i - 1/3}{n + 1/3}\) be the
empirical probability of instance \(i\). The normal distribution is
parametrised by location parameter \(\mu\) and scale parameter
\(\sigma\), both of which have to be estimated from the data. For
non-robust power transformations, \(\mu\) and \(\sigma\) are sample mean
and sample standard deviation, respectively. For robust power
transformations, we estimate \(\mu\) and \(\sigma\) as Huber M-estimates
of location and scale of the transformed feature
\(\phi^{\lambda, x_0, s} (\mathbf{X})\) (Huber 1981).

@@

Here \(w_i\) are weights, and \(\mu\) and \(\sigma\) are location and
scale parameters. For non-robust power transformations, all \(w_i = 1\).
The criterion is similar to the Cramér-von Mises test statistic Mises
(1928), aside from a additive scalar value and the introduction of
weights. This criterion, like the Anderson-Darling criterion, seeks to
minimise the squared Euclidean distance between the EDF and the normal
CDF. Unlike the Anderson-Darling criterion, this criterion weights all
instances equally.

>>>>

Many statistical and some machine learning methods assume normality of
the underlying data, e.g.~analysis of variance and linear discriminant
analysis. However, numerical features in datasets may strongly deviate
from normal distributions, e.g.~by being skewed. Power transformations
aim to stabilise variance and improve normality of such features
\citep{Bartlett1947-rx, Tukey1957-rt}. The two most commonly used
transformations are that of \citet{Box1964-mz} and \citet{Yeo2000-vw}.

@@

Real-world data may contain outliers, to which maximum likelihood
estimation can be sensitive. Their presence may lead to poor
transformations to normality, as shown in Figure
\ref{fig:decreased-normality}. As indicated by \citet{Raymaekers2024-zf},
 the general aim of power transformations should be to transform
non-outlier data to normality, i.e.~achieve \emph{central normality}. To
achieve this, they devised an iterative procedure to find a robust
estimate of the transformation parameter \(\lambda\). Briefly, this
process requires identifying outliers in the data and weighting such
instances during the optimisation process. \citet{Raymaekers2024-zf}
achieve this through weighted maximum likelihood estimation.
However, because this procedure iteratively estimates and updates
\(\lambda\), it can not be used here to simultaneously estimate
\(\lambda\), \(x_0\) and \(s\) for location- and scale-invariant power
transformations. Nonetheless, as a procedure, weighted MLE can be used
for estimating the transformation, shift and scale parameters.

Here, weighted maximum likelihood estimation is based on equations
\ref{eqn:box-cox-invariant-log-likelihood} and
\ref{eqn:yeo-johnson-invariant-log-likelihood}. Compared to \citet{Raymaekers2024-zf},
these log-likelihood functions include additional
terms to accommodate estimation of \(x_0\) and \(s\). The weighted
location- and scale-invariant Box-Cox log-likelihood function is:

@@

\begin{itemize}
\item
  A step function, with \(\delta_1 \geq 0\) as threshold parameter:
  \begin{equation*}
  w_i =
  \begin{cases}
  1 & \text{if } \left| \dot{x}_i \right| \leq \delta_1\\
  0 & \text{if } \left| \dot{x}_i \right| > \delta_1
  \end{cases}
  \end{equation*}
\item
  A triangle function (or generalised Huber weight), with
  \(\delta_1 \geq 0\) and \(\delta_2 \geq \delta_1\) as threshold
  parameters: \begin{equation*}
  w_i =
  \begin{cases}
  1 & \text{if } \left| \dot{x}_i \right| < \delta_1\\
  1 - \frac{\left| \dot{x}_i \right| - \delta_1}{\delta_2 - \delta_1} & \text{if } \delta_1 \leq \left| \dot{x}_i \right| \leq \delta_2 \\
  0 & \text{if } \left| \dot{x}_i \right| > \delta_2
  \end{cases}
  \end{equation*}
\item
  A tapered cosine function \citep{Tukey1967-eb}, with \(\delta_1 \geq 0\)
  and \(\delta_2 \geq \delta_1\) as threshold parameters:
  \begin{equation*}
  w_i =
  \begin{cases}
  1 & \text{if } \left| \dot{x}_i \right| < \delta_1\\
  0.5 + 0.5 \cos\left(\pi \frac{\left| \dot{x}_i \right| - \delta_1}{\delta_2 - \delta_1} \right) & \text{if } \delta_1 \leq \left| \dot{x}_i \right| \leq \delta_2 \\
  0 & \text{if } \left| \dot{x}_i \right| > \delta_2
  \end{cases}
  \end{equation*}
\end{itemize}

@@

We require a quantile function (or an approximation thereof) to draw
random values from an asymmetric generalised normal distribution using
inverse transform sampling. Gijbels et al.~derived the quantile function
\(F_{\alpha}^{-1}(p)\), which incorporates the quantile function of the
symmetric generalised normal distribution derived by Griffin \citep{Gijbels2019-te}:

@@

Power transformations aim to transform features to a normal
distribution. However, this may not always be successful or possible.
Deviations from normality can be detected by normality tests, such as
the Shapiro-Wilk test  \citep{Gijbels2019-te}. In practice, normality
tests may be too stringent with large sample sizes, outliers, or both.
Here we develop a test for central normality.

@@

Let the residual error for each element of $\mathbf{X}$ be $r_i =\left| \frac{x_i - \mu_M}{\sigma_M} - F^{-1}_{\mathcal{N}}(p_i) \right|$, 
with $\mu_M$ and $\sigma_M$ robust Huber M-estimates of location and scale of $\mathbf{X}$ \citep{Huber1981-su}, 
and $F^{-1}_{\mathcal{N}}$ the quantile function of the normal distribution $\mathcal{N}(0, 1)$.

@@

We used simulated data to assess invariance to location and scale of the
proposed power transformations, to develop the empirical central
normality test, and to determine weighting for robust transformations.
The \(\lambda\) parameter for conventional power transformations (Eqn.
\ref{eqn:box-cox-original} and \ref{eqn:yeo-johnson-original}), as well
as \(\lambda\), \(x_0\) and \(s\) parameters for location- and
scale-invariant power transformations (Eqn. \ref{eqn:box-cox-invariant}
and \ref{eqn:yeo-johnson-invariant}) were estimated using the BOBYQA
algorithm for derivative-free bound constraint optimisation \citep{Powell2009-zb} 
through maximum likelihood estimation. The required algorithms
were implemented in the \texttt{power.transform} R software package
\citep{Zwanenburg2024-kq} (version 1.0.1). Of note, the
\texttt{power.transform} package shifts feature values into the positive
domain if negative or zero values are present for Box-Cox power
transformations.

@@

Outlier values were then drawn to randomly replace 10 percent of the
elements of \(\mathbf{X}_i\). Outlier values were set according to \citet{Tukey1977-xm},
as follows. Let \(x^{*} \sim U\left(-2, 2\right)\).
Then the corresponding outlier value was:

@@

In addition to no power transformation and location- and scale-invariant
power transformations, conventional and Raymaekers and Rousseeuw's
robust adaptation \citep{Raymaekers2024-zf} were assessed. For the
latter two methods, normalisation before standardisation using was
additionally assessed using the following two methods:

@@

A common feature in health-related datasets is age. Here we use data on
228 patients with lung cancer that was collected and published by
Loprinzi et al. \citep{Loprinzi1994-cd}. The age in the cohort was
\(62.4 \pm 9.1\) (mean ± standard deviation) years. Applying
conventional and invariant Yeo-Johnson transformations to patient age
yielded the following results, see Figure
\ref{fig:experimental-results-invariance}: no transformation
(\(\sum r_i = 16.5\), \(p = 0.69\)); conventional transformation
(\(\lambda = 2.0\), \(\sum r_i = 11.5\), \(\mu_{YJ} = 1.8 \cdot 10^3\),
\(\sigma_{YJ} = 0.5 \cdot 10^3\), \(p = 0.96\)); Raymaekers and
Rousseeuw's robust adaptation (\(\lambda = 2.0\), \(\sum r_i = 11.5\),
\(\mu_{YJ} = 1.8 \cdot 10^3\), \(\sigma_{YJ} = 0.5 \cdot 10^3\),
\(p = 0.96\)); location- and scale-invariant transformation
(\(\lambda = 1.3\), \(\sum r_i = 8.8\), \(\mu_{YJ} = -1.2\),
\(\sigma_{YJ} = 1.1\), \(p = 0.98\)); and robust location- and
scale-invariant transformation (\(\lambda = 1.3\), \(\sum r_i = 9.3\),
\(\mu_{YJ} = -1.0\), \(\sigma_{YJ} = 1.1\), \(p = 0.93\)).

@@

Gorman, Williams and Fraser recorded body mass (in grams) of 342
penguins of three different species \citep{Gorman2014-eo}.
The body mass was \((4.2 \pm 0.8) \cdot 10^3\) (mean ± standard
deviation) grams, and not centrally normal (\(p < 0.001\)). Applying
conventional and invariant Yeo-Johnson transformations to body mass
yielded the following results, see Figure
\ref{fig:experimental-results-invariance}: no transformation (residual
sum \(\sum r_i = 48.0\), \(p < 0.001\)); conventional transformation
(\(\lambda = -0.5\), \(\sum r_i = 32.2\), \(\mu_{YJ} = 2.1\),
\(\sigma_{YJ} = 4 \cdot 10^{-3}\), \(p = 0.10\)); Raymaekers and
Rousseeuw's robust adaptation (\(\lambda = -0.5\), \(\sum r_i = 32.2\),
\(\mu_{YJ} = 2.1\), \(\sigma_{YJ} = 4 \cdot 10^{-3}\), \(p = 0.10\));
location- and scale-invariant transformation (\(\lambda = 0.5\),
\(\sum r_i = 26.8\), \(\mu_{YJ} = 0.9\), \(\sigma_{YJ} = 0.9\),
\(p = 0.28\)); and robust location- and scale-invariant transformation
(\(\lambda = 0.3\), \(\sum r_i = 22.0\), \(\mu_{YJ} = 0.7\),
\(\sigma_{YJ} = 0.9\), \(p = 0.69\)).

@@

Geospatial datasets usually contain coordinates. The Ames housing
dataset contains data on 2930 properties that were sold between 2006 and
2010 \citep{De-Cock2011-jf}, including their geospatial coordinates. The
latitude was \(42.03 \pm 0.02)\) (mean ± standard deviation). Applying
conventional and invariant Yeo-Johnson transformations to latitude
yielded the following results, see Figure
\ref{fig:experimental-results-invariance}: no transformation (residual
sum \(\sum r_i = 328\), \(p < 0.001\)); conventional transformation
(\(\lambda = 62.1\), \(\sum r_i = 319\),
\(\mu_{YJ} = 4.8 \cdot 10^{99}\), \(\sigma_{YJ} = 0.1 \cdot 10^{99}\),
\(p < 0.001\)); Raymaekers and Rousseeuw's robust adaptation
(\(\lambda = 95.4\), \(\sum r_i = 315\),
\(\mu_{YJ} = 6.4 \cdot 10^{153}\), \(\sigma_{YJ} = 0.3 \cdot 10^{153}\),
\(p < 0.001\)); location- and scale-invariant transformation
(\(\lambda = 1.5\), \(\sum r_i = 326\), \(\mu_{YJ} = -1.2\),
\(\sigma_{YJ} = 0.8\), \(p < 0.001\)); and robust location- and
scale-invariant transformation (\(\lambda = 1.1\), \(\sum r_i = 308\),
\(\mu_{YJ} = -1.3\), \(\sigma_{YJ} = 1.2\), \(p < 0.001\)).

@@

The Top Gear dataset contains data on 297 vehicles that appeared on the
BBC television show \emph{Top Gear} \citep{Alfons2021-kc}. Within this dataset,
the fuel consumption feature contains outliers due to highly
fuel-efficient vehicles. Applying conventional and invariant Yeo-Johnson
transformations to the fuel consumption feature yielded the following
results, see Figure \ref{fig:experimental-results-outlier-robustness}:
no transformation (residual sum \(\sum r_i = 54\), \(p=0.72\));
conventional transformation (\(\lambda = -0.1\), \(\sum r_i = 55\),
\(\mu_{YJ} = 3.0\), \(\sigma_{YJ} = 0.3\), \(p < 0.001\)); Raymaekers
and Rousseeuw's robust adaptation (\(\lambda = 0.8\), \(\sum r_i = 48\),
\(\mu_{YJ} = 29\), \(\sigma_{YJ} = 15\), \(p=0.37\)); location- and
scale-invariant transformation (\(\lambda = -1.3\), \(\sum r_i = 44\),
\(\mu_{YJ} = 0.5\), \(\sigma_{YJ} = 0.1\), \(p < 0.001\)); and robust
location- and scale-invariant transformation (\(\lambda = 1.0\),
\(\sum r_i = 56\), \(\mu_{YJ} = 1.7\), \(\sigma_{YJ} = 1.0\),
\(p=0.76\)).

@@

The ischemic stroke dataset contains historic data from 126 patients
with risk at ischemic stroke \citep{Kuhn2019-kt}. These patients
underwent Computed Tomography Angiography to characterize the carotid
artery blockages. Angiography imaging was then assessed, and various
characteristics related to the blood vessels and the disease are
measured. The maximum arterial wall thickness feature contains several
instances with outlier values. Applying conventional and invariant
Yeo-Johnson transformations to this feature yielded the following
results, see Figure \ref{fig:experimental-results-outlier-robustness}:
no transformation (residual sum \(\sum r_i = 110\), \(p=0.83\));
conventional transformation (\(\lambda = -0.7\), \(\sum r_i = 30\),
\(\mu_{YJ} = 1.0\), \(\sigma_{YJ} = 0.1\), \(p=0.003\)); Raymaekers and
Rousseeuw's robust adaptation (\(\lambda = 1.1\), \(\sum r_i = 136\),
\(\mu_{YJ} = 7.2\), \(\sigma_{YJ} = 14.3\), \(p=0.88\)); location- and
scale-invariant transformation (\(\lambda = 0.2\), \(\sum r_i = 12\),
\(\mu_{YJ} = -11.8\), \(\sigma_{YJ} = 6.9\), \(p=0.15\)); and robust
location- and scale-invariant transformation (\(\lambda = -0.3\),
\(\sum r_i = 30\), \(\mu_{YJ} = 0.8\), \(\sigma_{YJ} = 0.2\),
\(p=0.18\)).

@@

We used 231 datasets containing at least one numeric feature from the
Penn Machine Learning Benchmarks collection \citep{Romano2022-gq}. In
this collection, 114 datasets correspond to regression tasks and 117
datasets to classification tasks. Using the familiar auto-machine
learning library \citep{Zwanenburg2021-so} (version 1.5.0), each
dataset was used to train a model for each of 32 process configurations.
Each process configuration specifies the learner (generalised linear
model, L1-regularised linear models (Lasso), gradient boosted linear
model, or random forest), transformation method (none, conventional
Yeo-Johnson, robust invariant Yeo-Johnson, robust invariant Yeo-Johnson
with empirical central normality test (rejecting transformations with
\(p \leq 0.01\)), and normalisation method (none,
\(z\)-standardisation), yielding 32 distinct configurations. Before each
experiment, each dataset was randomly split into a training (70\%) and
holdout test (30\%) set five times. Thus, a total of 36960 models were
created. Each model was then evaluated using the holdout test set using
one of two metrics, i.e.~the root relative squared error (RRSE) for
regression tasks and the area under the receiver operating
characteristic curve (AUC) for classification tasks.

@@

To estimate the marginal effects of process parameters, including
transformation method, we first fit a regression random forest (ranger
package \citep{Wright2017-rf} version 0.16.0): 2000 trees, node size
2, other hyperparameters default) with process parameters and task
difficulty as predictors and normalised rank as response variable. The
estimated marginal effects are shown in Figure
\ref{fig:marginal-effect-plot}. On the scale of normalised ranks
(\([0.0, 1.0]\)), only the random forest performed better than the
average of all learners (rank difference \(0.173\)). The marginal
improvement in performance from z-standardisation was \(0.011\).
Transformation methods had the following marginal effects: \(-0.001\)
for using conventional Yeo-Johnson transformation instead of no
transformation; \(-0.012\) for using robust invariant Yeo-Johnson
transformation instead of no transformation; \(-0.012\) for using robust
invariant Yeo-Johnson transformation with empirical central normality
test instead of no transformation; \(-0.012\) for using robust invariant
Yeo-Johnson transformation instead of conventional Yeo-Johnson
transformation; and \(-0.012\) for using using robust invariant
Yeo-Johnson transformation with empirical central normality test instead
of conventional Yeo-Johnson transformation.

@@

In their work on power transformation, Box and Cox already mention
transformation with a shift parameter, but preferred the version in Eq.
\ref{eqn:box-cox-original} for the theoretical analysis in their paper
\citep{Box1964-mz}, which subsequently became the convention. Yeo and
Johnson's power transformation lacks a shift parameter altogether \citep{Yeo2000-vw}.
We showed that these power transformations are
sensitive to location and scale of data distributions. To mitigate this
issue, we defined location- and scale-invariant variants of the Box-Cox
and Yeo-Johnson transformations. We furthermore assessed methods for
making these transformations robust to outliers, and devised an
empirical test for central normality.

@@

We introduced an empirical test for central normality to assess whether
sequences deviate from normality in a way that might require closer
inspection prior to further processing. The empirical test for central
normality differs from other tests for normality, such as the
Shapiro-Wilk test \citep{Shapiro1965-zd}, as it assesses normality of
the central portion of a feature, instead of the entire feature.
Compared to the central normality test and the Shapiro-Wilk test, the
empirical central normality test remains consistent in the presence of
outliers, although the former tests are more powerful when outliers are
absent.

@@

Below we will focus on criteria based on the empirical density function
and those based on skewness and kurtosis of the transformed featured.
Other potential criteria, such as the Shapiro-Wilk test statistic
\citet{Raymaekers2024-zf} are not investigated here. In the case of the
Shapiro-Wilk test statistic this is because of lack of scalability to
features with many (\(> 5000\)) instances, and because adapting the test
statistic to include weights is not straightforward.

@@

The first class of criteria is based on the empirical distribution
function (EDF). Transformation parameters are then fit through
minimisation of the distance between the empirical distribution function
\(F_{\epsilon}\) and the cumulative density function (CDF) of the normal
distribution \(F_{\mathcal{N}}\). Let
\(F_{\epsilon}\left(x_i \right) = \frac{i - 1/3}{n + 1/3}\) be the
empirical probability of instance \(i\). The normal distribution is
parametrised by location parameter \(\mu\) and scale parameter
\(\sigma\), both of which have to be estimated from the data. For
non-robust power transformations, \(\mu\) and \(\sigma\) are sample mean
and sample standard deviation, respectively. For robust power
transformations, we estimate \(\mu\) and \(\sigma\) as Huber M-estimates
of location and scale of the transformed feature
\(\phi^{\lambda, x_0, s} (\mathbf{X})\) \citep{Huber1981-su}.

@@

Here \(w_i\) are weights, and \(\mu\) and \(\sigma\) are location and
scale parameters. For non-robust power transformations, all \(w_i = 1\).
The criterion is similar to the Cramér-von Mises test statistic 
\citep{Cramer1928-rc, Von_Mises1928-ef}, aside from a additive scalar value and the introduction of
weights. This criterion, like the Anderson-Darling criterion, seeks to
minimise the squared Euclidean distance between the EDF and the normal
CDF. Unlike the Anderson-Darling criterion, this criterion weights all
instances equally.
